---
title: "Tidyverse"
output:
  pdf_document: default
  html_document: default
date: "2023-11-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Tidyverse

The Tidyverse is an ecosystem of compatible packages that share the same programming philosophy, centered around the concept of "tidy" data, i.e. data that is formatted in a specific way to make it easy to analyze and visualize.

### Pipes

There are several types of pipe operators provided by the `magrittr` package:

+ `%>%`: takes the output from one function/operation and passes it as an element (by default, the first) to the next function, enabling a flow from left-to-right (rather than from inner functions to outer functions). This is the most commonly used pipe operator in R
+ `%<>%`: performs the piped operations and stores the result via assignment (must be the first pipe)
+ `%T>%`: allows the storage of the partial results of a pipe, which is especially useful if the final results are a side-effect such as plotting or printing
+ `%$%`: makes the names of the object on the left-hand side of the pipe available in the function call on the right-hand side

### `ggplot2`

Similarly to how pipes enable the flow of objects (often data) from left to right, the data visualization package `ggplot2` enables the addition of graphical layers to plots from left to right. In `ggplot2`, each additional layer is added using a `+` sign, allowing the specification of data, aesthetics (e.g. color, line-type, etc.), and geometries (e.g. points, lines, etc.) at each stage. This "Grammar of Graphics" makes visualizations highly customizable, far more than the plotting functions in base R.

## Energy example

For the remainder of this portfolio, I will use tidyverse packages to manipulate and visualize some of the data available in the [AHSRAE - Great Energy Predictor III](https://www.kaggle.com/competitions/ashrae-energy-prediction/overview) Kaggle dataset.

### Import data

I start by using the `readr` package to load in several of the datasets from Kaggle:

```{r}
library(readr)

# Import train data
train <- read_csv("../data/train.csv")
building_metadata <- read_csv("../data/building_metadata.csv")
```

We can view the first few rows of each dataset:

```{r}
head(train)
head(building_metadata)
```

### Join data

We can see that both datasets contain a `building_id` column, which is the primary key for `building_metadata` as well as a foreign key linking the two datasets. We can use functions from `dplyr` to join the datasets:

```{r}
library(dplyr)

train_building_metadata <- train %>% 
  left_join(building_metadata, by = "building_id")
head(train_building_metadata)
```

### Summarize data

We can use other `dplyr` functionality to calculate some aggregate values based on the metadata:

```{r}
# Calculate summary values for meter readings
# For each building
agg_reading_per_building <- train_building_metadata %>% 
  group_by(building_id) %>% 
  summarise(
    mean_meter_reading = mean(meter_reading),
    median_meter_reading = median(meter_reading),
    max_meter_reading = max(meter_reading)
  )
head(agg_reading_per_building)

# Do the same grouped by site
agg_reading_per_site <- train_building_metadata %>% 
  group_by(site_id) %>% 
  summarise(
    mean_meter_reading = mean(meter_reading),
    median_meter_reading = median(meter_reading),
    max_meter_reading = max(meter_reading)
  )
agg_reading_per_site

# Do the same grouped by the building use
agg_reading_per_use <- train_building_metadata %>% 
  group_by(primary_use) %>% 
  summarise(
    mean_meter_reading = mean(meter_reading),
    median_meter_reading = median(meter_reading),
    max_meter_reading = max(meter_reading)
  )
agg_reading_per_use
```

### Visualize data

It appears that there are large differences in meter readings based on the primary use of the building. We can visualize the distribution of meter readings by primary use using `ggplot2`:

```{r}
library(ggplot2)
library(forcats) # to allow us to reorder the boxplot

train_building_metadata %>% 
  # Fix situations where meter reading was 0 for the log-transform
  mutate(meter_reading = ifelse(meter_reading == 0,
                                meter_reading + 0.000000001,
                                meter_reading)) %>% 
  # Create variable of log-transformed meter reading for viz purposes
  mutate(log_meter_reading = log(meter_reading)) %>% 
  ggplot(aes(
    # Order the boxplot by descending median log meter reading
    x = fct_reorder(primary_use, log_meter_reading, .desc = TRUE),
    y = log_meter_reading
  )
  ) +
  geom_boxplot() +
  labs(x = "Building use",
       y = "Log meter reading") +
  # Rotate the x-axis labels so we can read them
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Note that creating this plot is quite slow given the size of our data, but it allows us to illustrate some of the nice properties of `ggplot2`.

Aggregating all meter readings for buildings of a particular use may help us see overall patterns, but meter usage is likely to fluctuate substantially over time. We can examine average daily meter readings per building use over time:

```{r}
library(lubridate) # for date manipulations

# Create a dataframe of the average daily meter reading by building use
avg_daily_meter <- train_building_metadata %>% 
  mutate(meter_date = as_date(timestamp)) %>% 
  group_by(primary_use,
           meter_date) %>% 
  summarise(med_reading = median(meter_reading))

ggplot(data = avg_daily_meter,
       aes(x = meter_date,
           y = med_reading,
           color = primary_use)) +
  geom_line() +
  labs(x = "Date", y = "Median meter reading", color = "Building use")
```

We can also use `facet_wrap()` to plot each building use in its own plot arranged in a grid:

```{r}
ggplot(data = avg_daily_meter,
       aes(x = meter_date,
           y = med_reading,
           color = primary_use)) +
  geom_line() +
  facet_wrap(~ primary_use) +
  labs(x = "Date", y = "Median meter reading", color = "Building use") +
  # Decrease the number of x-axis breaks to 3
  scale_x_continuous(breaks = seq(min(avg_daily_meter$meter_date), max(avg_daily_meter$meter_date), length.out = 3)) +
  theme(
    # Remove the color legend which takes up a ton of space
    legend.position = "none",
    # Rotate the x-axis labels so we can read them
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )
```


