---
title: "Numerical integration"
output: pdf_document
author: "Cecina Babich Morrow"
date: "2023-11-07"
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Integration

Many applications in statistics necessitate the calculation of integrals. Unfortunately many integrals cannot be computed in closed form, necessitating approximation techniques.

## Quadrature

"Quadrature rules" refer to the process of approximating integrals, specifically in a deterministic way. Polynomial interpolation is at the heart of quadrature. To use quadrature, we'll start by numerically approximating the definite integral of a function over a finite interval using a polynomial function, since polynomials can be integrated exactly.

### Gaussian quadrature

Gaussian quadrature attempts to approximate an integral as a weighted summation. There are many types of Gaussian quadrature, so this portfolio will focus on Gauss-Legendre quadrature. It aims to find an accurate approximation of the integral by selecting specific points (nodes) and associated weights for the integration. In the Gauss-Legendre quadrature, these nodes and weights are chosen in such a way that the method provides exact results for polynomials up to a certain degree. The nodes are typically roots of orthogonal polynomials, and the weights are determined to ensure optimal accuracy. The Gauss-Legendre quadrature is particularly effective for integrating functions over a finite interval.

Gauss-Legendre quadrature takes the form
$$\int_{-1}^1 f(x) dx \approx \sum_{i = 1}^n w_i f(x_i)$$
for a set of $n$ sample points and $w_i$ weights, where $x_i$ are the roots of the $n$th Legendre polynomials. This rule can provide exact integration of degree $2n - 1$ polynomials.

```{r}
library(statmod)
# Calculate nodes and weights where n = 3
gauss_3 <- gauss.quad(n = 3, kind = "legendre")
# where n = 8
gauss_8 <- gauss.quad(n = 8, kind = "legendre")
```

```{r}
example_poly <- function(x) 1 + x + x^2 + x^3 + x^4 + x^5
curve(example_poly, from = -10, to = 10)
```

```{r}
gauss_legendre_approx <- function(f, gauss_quad_output) {
  weights <- gauss_quad_output$weights
  nodes <- gauss_quad_output$nodes
  approx <- sum(weights * f(nodes))
  return(approx)
}

gauss_legendre_approx(example_poly, gauss_3)
```

This is a very good approximation of the true integral (which is $\frac{46}{15} \approx 3.0666667$). Let's try a higher order polynomial:

```{r}
higher_order_poly <- function(x) 1 + x + x^2 + x^3 + x^4 + x^5 + x^8 + 11*x^10
curve(higher_order_poly, from = -10, to = 10)

gauss_legendre_approx(higher_order_poly, gauss_3)
gauss_legendre_approx(higher_order_poly, gauss_8)
```

As the number of nodes increases, we get a better approximation of the true integral ($\frac{238}{45} \approx 5.2889$).









