---
title: "Matrix operations"
output: pdf_document
author: "Cecina Babich Morrow"
date: "2023-11-28"
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrices

A matrix in R is a two-dimensional data structure consisting of columns and rows. They can be constructed using the `matrix()` function, with entries specified by column by default.

### Numerical stability

R has an error of order $10^{-16}$. The `all.equal()` function checks for "near equality" up to a given tolerance, set to $1.5 \times 10^{-8}$ by default, which we can compare to `==`:

```{r}
one_plus <- 1 + 1e-15
one <- 1
one_plus
one
one_plus == one
all.equal(one_plus, one)
```

R can also store integers as a "long" type, as compared to the numeric type:

```{r}
class(1)
class(1L)
```


## Dense matrices

In the `Matrix` package, `dgeMatrix` is the S4 class for dense matrices.

```{r}
library(Matrix)
dense_matrix <- Matrix(c(1, 2, 4, 3, 5, 8, 7, 7, 7), nrow=3, ncol=3)
class(dense_matrix)
```

The following code demonstrates some basic matrix operations for dense matrices:

```{r}
# Invert a matrix:
solve(dense_matrix)
# Find the rank of a matrix:
rankMatrix(dense_matrix)
# Give the reciprocal condition number (reciprocal of product of norm of matrix and norm of inverse):
rcond(dense_matrix)
1 / (norm(dense_matrix) * norm(solve(dense_matrix)))
```

## Sparse matrices

Sparse matrices are matrices where most entries are 0. They are typically stored as class `dgCMatrix`, which are stored in a compressed, sparse, column-oriented format to save memory. 

```{r}
set.seed(17)
nrows <- 1000
ncols <- 1000
vals <- sample(x=c(0, 1, 2), prob=c(0.98, 0.01, 0.01), size=nrows*ncols, replace=TRUE)
sparse_matrix <- Matrix(vals, nrow=nrows, ncol=ncols, sparse=TRUE)
class(sparse_matrix)
```

The following code demonstrates some basic matrix operations for sparse matrices:

```{r}
# Invert a matrix:
sparse_inverse <- solve(sparse_matrix)
class(sparse_inverse) # still dgCMatrix
# Multiply by a vector:
class(sparse_inverse %*% c(1, rep(0, 998), 1)) # dgeMatrix
# Multiply by that vector represented as a sparse matrix:
class(sparse_inverse %*% Matrix(c(1, rep(0, 998), 1), nrow=1000, ncol=1, sparse=TRUE))
```

### Dependency graphs

We can use symmetric sparse matrices as adjacency graphs for graphs:

```{r}
library(igraph)
sym_sparse <- rsparsematrix(nrow = 10,
                            ncol = 10,
                            density = 0.1,
                            symmetric = TRUE)
adj_graph <- graph_from_adjacency_matrix(sym_sparse, weighted = TRUE)
plot(adj_graph)
```

